---
title: "Lab 7"
subtitle: "The Color of Drinking Water"
toc: true
editor: visual
format:
  html:
    other-links:
      - text: Download Source
        href: lab7.qmd
      - text: Drinking Water Data
        href: data/drinking_water.csv
editor_options: 
  chunk_output_type: inline
---

## Learning objectives

In today's lab you will...

1.  Fit **logistic regression** models to binary (0/1, no/yes) outcomes
2.  Interpret **coefficients**
3.  **Visualize** predictions

Complete sections **Logistic model notation and R code** and **Read and explore the data** before class.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
theme_set(theme_classic(12))
set.seed(123)

```

## Logistic model notation and R code

In statistical notation, logistic models take the form:

$$
\begin{align}
\text{BinaryOutcome} &\sim Binomial(1, p) \\
logit(p) &= \beta_0 + \beta_1 \text{Predictor}
\end{align}
$$

In R, you can fit a logistic model like this:

``` r
logistic_model <- glm(binary_outcome ~ predictor,
                      data = my_data,
                      family = binomial(link = "logit"))
```

**In both the notation and code above, identify the:**

-   **Response family**

-   **Link function**

-   **Response**

-   **Predictor**

## Read and explore the data

**Read the dataset in drinking_water.csv.**

```{r}
#| label: read-data

drinking_water <- read_csv('~/Desktop/MEDS/EDS-222/eds-222-labs/week7/drinking_water.csv')
head(drinking_water)
```

**Describe what you think each of the columns represents.**

| Column              | Meaning                              |
|---------------------|--------------------------------------|
| `pwsid`             | Public water system ID               |
| `pwsname`           | Public water system name             |
| `pctpov`            | Percent of local district in poverty |
| `pcthisp`           | Percent of local district hispanic   |
| `health`            |                                      |
| `medianincomehouse` | Median household income of district  |
| `violation`         | Water Violation                      |
| `pcthisp_bin`       | Percent Hispanic Bin                 |
| `pctpov_bin`        | Percent Poverty Bin                  |

**Create three figures summarizing the relationships between race, class, and SDWA violations.**

Figure 1: A scatter plot showing the *proportion* of districts with SDWA violations in each *percent Hispanic bin*.

Figure 2: A scatter plot showing the *proportion* of districts with SDWA violations in each *percent below-poverty bin*.

Figure 3: A scatter plot showing the correlation between *percent Hispanic* and *percent below-poverty* at the district level.

```{r}
#| label: fig-explore

water |> 
  group_by(pcthisp_bin) |> 
  summarise(prop_violation = mean(violation)) |> 
  ggplot(aes(pcthisp_bin, prop_violation)) +
  geom_point()


```

## Fit and interpret model

Our question is:

*How are SDWA violations distributed among communities by class and race?*

Let's begin by investigating class in isolation. **Fit a logistic model to see how `violation` responds to `pctpov`**.

```{r}
#| label: fit-model-class

class_model <- glm(violation ~ pctpov, # formula: response ~ predictor(s)
                   data = drinking_water, # data remains unhanged
                   family = binomial(link = "logit")) # Family and link function

```

Just as we did with linear regression, you can inspect coefficient estimates, p-values, and confidence intervals using `summary()` and `confint()`.

**What is the best estimate for the % Below Poverty coefficient?**

```{r}
coef(class_model)["pctpov"]
```

**Is the coefficient significantly different than 0? Assume critical threshold of 0.05.**

```{r}
# It is not significantly different than 0 meaning it is highly unlikely that a relationship occured just by chance
summary(class_model)$coefficient["pctpov", 4]
```

**What is the 95% CI for the coefficient?**

```{r}
confint(class_model)["pctpov", ]
```

**Based on the sign, what is the relationship between class and SDWA violations?**

Sign is positive, so small positive association between the two

## Make predictions

Interpreting the coefficients directly is pretty difficult with logistic regression! What is the difference between a coefficient of 0.1 and 0.2? Rather than interpreting the coefficient directly, it's often easier to compare predictions.

Once you have a model, you can make predictions using `predict()`. When calling `predict()`, the following parameters are important:

-   `object` The model you want to generate predictions from

-   `newdata` A data frame of predictors your want predictions for

-   `type` Do you want predictions on the *link* or *response* scale? By default, `predict()` shows the value of $logit(p) = \beta_0 + \beta_1 x$ (i.e., link-scale). To get the predicted *probability* (i.e., response-scale) set `type` to "response".

-   `se.fit` If `se.fit` is `TRUE`, `predict()` will calculate the confidence interval of the response on the link scale. You can use that to get the confidence interval of $p$. That will make a ribbon, similar to the confidence interval of $\mu$ you saw in linear regression. Gives you the relationship between X and logit(p).

Here's how to generate predictions (with CI) for a single predictor.

::: callout-tip
I use `expand_grid()` instead of `tibble()` when making the data frame of predictors because it gives you every combination of the values in the columns. That will be important when you have more than one predictor!
:::

```{r}
#| label: predict-class-model

# p = percent prediction of a water violation ocuring in an area of pctpov

# A grid of predictors(s) to generate predicitons for
pred_grid <- expand_grid(pctpov = 0:100)
# Generate predictions
class_model_pred <- pred_grid |>
  mutate(p = predict(object = class_model,
                     newdata = pred_grid,
                     type = "response"))

```

```{r}
#| label: plot-class-best-estimate

ggplot(class_model_pred, aes(x = pctpov, y = p)) +
  geom_line() +
  ylim(0,1)

```

That was just the best estimate. The following code adds the 95% CI for $p$. It uses the *inverse link function* to go from link-scale to response-scale.

::: callout-caution
When constructing the 95% CI for $p$, one thing remains the same and one thing has changed from constructing the 95% CI for $\mu$.

**What remains the same:** the sampling distribution is normal, so once you have the standard error you can use `qnorm()` to get the 2.5% and 97.5% quantiles.

**What's different:** the sampling distribution exists on the *link scale*. In other words, the standard error is $SE(logit(p))$ rather than $SE(p)$. So you have to get the best estimate and standard error on the link scale, then invert the link function to get back to $p$.
:::

```{r}
#| label: predict-class-model-ci

class_model_se <- predict(object = class_model,
                          newdata = pred_grid,
                          type = "link", # Stay in link space!
                          se.fit = TRUE) # Get the SE of the fit

# Go from link space to response space
linkinv <- family(class_model)$linkinv # Function to go from link to response

class_model_pred <- pred_grid |> 
  mutate(
    # Get logit (p)
    logit_p = class_model_se$fit,
    # 95% CI in link space
    logit_p_se = class_model_se$se.fit,
    logit_p_lwr = qnorm(0.025, mean = logit_p, sd = logit_p_se),
    logit_p_upr = qnorm(0.975, mean = logit_p, sd = logit_p_se),
    # Undo the link function usuing linkinv
    p = linkinv(logit_p),
    p_lwr = linkinv(logit_p_lwr),
    p_upr = linkinv(logit_p_upr)
  )


# Link space is where we can do meaningful math, response space is where we can interpret it becuase now we have the probability, not just the odds of the probability.

```

And here's the fit with the 95% CI presented as the familiar ribbon.

```{r}
#| label: plot-class-ci

ggplot(class_model_pred, aes(x = pctpov, y = p)) +
  geom_ribbon(aes(ymin = p_lwr, ymax = p_upr), alpha = 0.2) +
  geom_line()

# This tells us the average value of districts is probably down by 20. Because the CI increases so much as we go up. 
```

**What is the probability of a SDWA violation in a district where 5% of households are below the poverty line? When 50% are? Show the best estimate and 95% CI.**

```{r}
#| label: class-estimate-ci



```

## Your turn

Switzer and Teodoro demonstrated that the *interaction* between race and class is important for understanding the environmental justice dimensions of drinking water. You'll investigate that here through the following steps.

1.  Create a new model with an interaction term between `pctpov` and `pcthisp`.

    ```{r}
    int_model <- glm(violation ~ pctpov * pcthisp, # formula: response ~ predictor(s)
                       data = drinking_water, # data remains unhanged
                       family = binomial(link = "logit")) # Family and link function
    ```

2.  Use the summary and confidence intervals to answer the following questions:\
    **What is the 95% CI for the % Hispanic coefficient?**

    ```{r}

    summary(int_model)$coefficient["pcthisp", 4]
    summary(int_model)$coefficient["pctpov", 4]
    confint(int_model)['pcthisp', ]
    ```

    **Is the interaction between class and race significantly different than 0?**

    Not significantly different than 0 so low chance that a relationship occurred just by chance

3.  Create a new predictor data frame with combinations of *both* predictors

    ```{r}
    # Create predictor data frame grid with both predictors
    int_grid <- expand_grid(pctpov = 0:100,
                             pcthisp = 0:100)
    ```

4.  Generate predictions, including both the best estimate and 95% CI.

    ```{r}
    # Generate predictions
    int_model_pred <- int_grid |>
      mutate(p = predict(object = int_model,
                         newdata = int_grid,
                         type = "response"))
    ```

5.  Plot the predictions. In your plot, put `pcthisp` on the x-axis and probability of SDWA violations on the y-axis. Indicate `pctpov` using color and fill. To make this work, you'll have to filter your predictions to a few levels of `pctpov` (e.g., 0, 25, 50, and 75).

    ```{r}
    int_model_pred1 <- int_model_pred |> 
      filter(pctpov %in% c(0, 25, 50, 75))


    ggplot(int_model_pred1, aes(x = pcthisp, y = p, group = pctpov, color = pctpov)) +
      geom_line() +
      ylim(0,1)
    ```

```{r}
#| label: your-turn



int_model <- glm(violation ~ pctpov * pcthisp, # formula: response ~ predictor(s)
                   data = drinking_water, # data remains unhanged
                   family = binomial(link = "logit")) # Family and link function

summary(int_model)$coefficient["pcthisp", 4]
summary(int_model)$coefficient["pctpov", 4]
confint(int_model)['pcthisp', ]

# Generate predictions
int_model_pred <- int_grid |>
  mutate(p = predict(object = int_model,
                     newdata = int_grid,
                     type = "response"))
# Create predictor data frame grid with both predictors
int_grid <- expand_grid(pctpov = 0:100,
                         pcthisp = 0:100)

int_model_pred1 <- int_model_pred |> 
  filter(pctpov %in% c(0, 25, 50, 75))


ggplot(int_model_pred1, aes(x = pcthisp, y = p, group = pctpov, color = pctpov)) +
  geom_line() +
  ylim(0,1)

int_model_se <- predict(object = int_model,
                          newdata = int_grid,
                          type = "link", # Stay in link space!
                          se.fit = TRUE) # Get the SE of the fit

# Go from link space to response space
linkinv <- family(int_model)$linkinv # Function to go from link to response

int_model_pred <- int_grid |> 
  mutate(
    # Get logit (p)
    logit_p_int = int_model_se$fit,
    # 95% CI in link space
    logit_p_se = int_model_se$se.fit,
    logit_p_lwr = qnorm(0.025, mean = logit_p_int, sd = logit_p_se),
    logit_p_upr = qnorm(0.975, mean = logit_p_int, sd = logit_p_se),
    # Undo the link function usuing linkinv
    p = linkinv(logit_p_int),
    p_lwr = linkinv(logit_p_lwr),
    p_upr = linkinv(logit_p_upr)
  )


ggplot(int_model_pred, aes(x = pcthisp, y = p, group = pctpov, color = pctpov)) +
  geom_ribbon(aes(ymin = p_lwr, ymax = p_upr), alpha = 0.2) +
  geom_line()
```

The coefficient for an interaction term in a logistic regression model is very difficult to interpret on its own! But that value has important consequences for the lives of millions of people. Predictions are usually more straightforward to interpret than coefficients. Use your figure and the outputs of `summary()` and `confint()` to reflect on the following prompt. Reply to the corresponding thread on Slack with your reflection to get 1 token.

**In plain language, what is the relationship between race, class, and SDWA violations? Which communities face the greatest risks associated with drinking water? How would you use the statistical outputs (model summary, prediction figure) to describe that risk quantitatively? How could those statistical outputs be misinterpreted?**

The graphs show that both race and class increase the probability of a SDWA violation exponentially. The communities that face the greatest risk with drinking water are communities with high percent poverty with a highly significant p-value. The p-value for percent his panic shows the same positive correlation but the confidence interval for percent Hispanic is actually entirely negative which could be misinterpreted to mean higher percent Hispanic communities are actually associated with less SDWA violations.
